<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An Introduction to the twNlme package}
-->


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"'
    , fig.align="left"
    , fig.width=4.6*1.2, fig.height=3.2*1.2, dev.args=list(pointsize=10)
    )
knit_hooks$set(spar = function(before, options, envir) {
    if (before){
        par( las=1 )                   #also y axis labels horizontal
        par(mar=c(2.0,3.5,0,0)+0.3 )  #margins
        par(tck=0.02 )                          #axe-tick length inside plots             
        par(mgp=c(1.1,0.2,0) )  #positioning of axis title, axis labels, axis
     }
})
```

# An Introduction to **twNlme** package #

## Introduction ##

The **twNlme** is an extension of the **nlme** package that 
* unifies the usage of gnls and nlme
* helps with calculating prediction uncertainty including that of fixed and random effects


## Tree biomass example ##

### The Data and the allometric model ###

Diameter `dbh` (cm) and stem biomass (kg) of trees of in the dataset `Wutzler08BeechStem` have been reported by different authors.

We have to deal with heteroscedastic errors. The variance increases with diameter and stem biomass.

Furthermore, trees have been measured at different locations and with slightly different methods.
Hence, there are groupings in the data, and the records of one study are not independent of each other. 

```{r , result='hide'}
library(twNlme)
```
```{r, spar=TRUE}
data(Wutzler08BeechStem)
head(Wutzler08BeechStem)
ds <- Wutzler08BeechStem[order(Wutzler08BeechStem$dbh),]
plot( stem~dbh, col=author, data=ds, xlab="dbh [cm]", ylab="")
mtext( "stem biomass [kg]", 2, 2.8, las=0)
legend( "topleft", inset=c(0.01,0.01), levels(ds$author), col=1:13, pch=1 )
```

The objective is to estimate stem biomass and of forests based on  measured tree diameters. 

The functional relationship can be described by an allometric equation: $y = \beta_0 d^{\beta_1}$. 
Where $y$ is the stem biomass, $d$ is the diameter, and $\beta_i$ are estimated model coefficients.

A log-transformation yields a linear form. And the plot reveals a similar slope between authors but differences in the intercept: $log(y) = log(\beta_0) +  {\beta_1} log(d)$.

```{r , spar=TRUE}
plot( log(stem)~log(dbh), col=author, data=subset(ds, dbh >= 7) )
```

### Ignore the grouping structure ###

First, the data is fitted with ignoring the grouping stucture, which leads to a severe underestimation of uncertainty of aggregated stem biomass predictions.

```{r , spar=TRUE}
lm1 <- lm( log(stem)~log(dbh), data=ds )
.start <- structure( c( exp(coef(lm1)[1]), coef(lm1)[2] ), names=c("b0","b1")) 

gm1 <- gnls( stem ~ b0 * dbh^b1 
    , data=ds 
    , params=c(b0+b1~1)
    , start=.start   
    , weights=varPower(form = ~fitted(.))
)
gm1

plot( stem~dbh, data=ds, xlab="dbh [cm]", ylab="")
mtext( "stem biomass [kg]", 2, 2.5, las=0)
lines( fitted(gm1) ~ dbh, data=ds, col="black", lwd=2 )

sdEps <- sqrt( varResidPower(gm1, pred=fitted(gm1)) )
gm1a <- attachVarPrep(gm1, fVarResidual=varResidPower)
sd <- varPredictNlmeGnls(gm1a, newdata=ds)[,"sdInd"] 
lines( fitted(gm1)+sd ~ dbh, data=ds, col="grey" )
lines( fitted(gm1)-sd ~ dbh, data=ds, col="grey" )
```
Note, the usage of function `varResidPower` to calculate uncertainty of the residuals that depends on the fitted values.

In addition to residual variance `varResid`, there is prediction uncertainty due to uncertainty in model coefficients: `varFix`. This component is, however, 3 orders of magnitudes smaller. Hence, there the prediction uncertainty of the population `sdPop` is much smaller than the prediction uncertainty of an individual `sdInd`.
  
```{r , spar=TRUE}
apply( varPredictNlmeGnls(gm1a, newdata=ds), 2, median )
```
Note, the usage of function `varPredictNlmeGnls` to calculate variance due to uncertainty in model coefficients.
It makes use of a Taylor approximation and needs to evaluate the derivatives of the statistical model at the predictor values. These derivative functions are symbolically derived before by calling function `attachVarPrep`.

Now, we use the model to calculate stem biomass and its uncertainty for n=1000 trees with a diameter of about 30cm.

```{r predictNoGroup, spar=TRUE, cache=TRUE}
n <- 1000      # sum over n trees
set.seed(0815)
dsNew <- data.frame(dbh=rnorm(n, mean=30, sd=2), author=ds$author[1])
yNew <- predict(gm1a, newdata=dsNew )
varNew <- varPredictNlmeGnls(gm1a, newdata=dsNew)
median( varNew[,"sdInd"]/yNew )       # on average each tree biomass with 33% uncertainty
yAgg <- sum(yNew)
# completely independent:
cvYAggRes <- sqrt( sum(varNew[,"varResid"]) ) /yAgg       # only 1% uncertainty ?
# indpendent but account for uncertainty in model coefficients
varSumComp <- varSumPredictNlmeGnls(gm1a, newdata=dsNew, retComponents = TRUE)
cvYAggInd <- varSumComp["sdPred"]/yAgg       # 3.1%
# dependent errors: add standard deviation
cvYAggDep <- sum( sqrt(varNew[,"varResid"]) ) /yAgg       # 33%: no decrease with increasing n
structure( c( cvYAggRes, cvYAggInd, cvYAggDep )*100, names=c("Residuals","Independent","Correlated"))
varSumComp/(n*10)
```

With aggregating over n=1000 trees, the residual uncertainty `varResid` declines and the uncertainty of model coefficients `varFix` becomes more important. Accounting for model coefficients raises the coefficient of variation from 1% to 3%.

However, we cannot treat the predictions as independent of each other.
To be conservative we assumme that errors add up and report a cv of 33%.

Note, how the aggregated uncertainty of model coefficients has been calculated by function 
`varSumPredictNlmeGnls`. The calculation involves the evaluation of the derivative functions for each compination of
predictors. Hence, it scales with $O(n^2)$ and takes some time for large n. 

### Accounting for the grouping with mixed models ###

Now we refit the model using random effects in coefficient $\beta_0$. I.e. we assume that it can vary between authors
around a common central value. Additionally, we assume that increase in residual variance with stem biomass can differ between authors. 

$$latex
y = ( \beta_0 + b_{0,k}) d^{\beta_1} + \epsilon 
$$

$$latex
b_{0,k} \sim\ \mathcal{N}(0,\,\sigma_k^2)
$$

```{r nlmeFit, spar=TRUE, cache=TRUE}
mm1 <- nlme( stem ~ b0 * dbh^b1 
        , data=ds 
        , fixed=c(b0+b1~1)
        , random=list(author=c(b0 ~1))
        #, random=list(author=c(b1 ~1))
        #, random=list(author=c(b0+b1 ~1))
        , start=.start   
        , weights=varPower(form = ~fitted(.)|author)
)
mm1
plot( stem~dbh, col=author, data=ds, xlab="dbh [cm]", ylab="")
mtext( "stem biomass [kg]", 2, 2.5, las=0)
yds <- predict(mm1, level=0) 
lines( yds ~ dbh, data=ds, col="black", lwd=2)

#authors <- levels(ds$author)
authors <- c("Heinsdorf","Cienciala","Duvigneaud2")
for( authorI in authors){
    dsNa <- ds; dsNa$author[] <- authorI
    try(lines( predict(mm1, level=1, newdata=dsNa) ~ dbh, data=dsNa, col=which(authorI == levels(ds$author)) ))
}
legend( "topleft", inset=c(0.01,0.01), legend=c(sort(authors),"Population"), col=c(which(levels(ds$author) %in% authors),"black"), lty=1, lwd=c(rep(1,length(authors)),2) )

#sd <- sqrt( varResidPower(gm1, pred=predict(mm1, level=0)) )
mm1a <- attachVarPrep( mm1, fVarResidual=varResidPower)     # calculate symbolic derivatives used in variance prediction
sd <-varPredictNlmeGnls(mm1a, newdata=ds)[,"sdInd"]     # variance components
lines( yds+sd ~ dbh, data=ds, col="grey" )
lines( yds-sd ~ dbh, data=ds, col="grey" )
```

The uncertainty of a single prediction did not change. However, the variance across the populations by the authors `varRan` is now about as large as the residual variance `varResid`:
```{r , spar=TRUE}
apply( varPredictNlmeGnls(mm1a, newdata=ds), 2, median )
```

This is significant, for the error propagation, because the uncertainty in model coefficients does not decrease with the number of measured trees. 

```{r predictGroup, spar=TRUE, cache=TRUE}
yNew <- predict(mm1, newdata=dsNew, level=0 )
yAgg <- sum(yNew)
(varSumCompM <- varSumPredictNlmeGnls(mm1a, newdata=dsNew, retComponents = TRUE))
(cvAgg <- varSumCompM["sdPred"] / yAgg )     # 18% uncertainty
```

When accounting for the groups, the uncertainty for the prediction of 1000 trees of an unknown group of 18% is much higher than the uncertainty of 3.3% estimated by ignoring the groups.
 
